---
title: "02ggplot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Gentle ggplot {#ggplot1}

## Preliminary steps

Load packages

```{r}
pacman::p_load(ggplot2, gapminder, tidyverse, car, plyr, ggthemes, devtools, stats, viridis, plotly)
```

Load the gapminder dataset

```{r, eval = FALSE}
gapminder <- gapminder

# What do the first 6 observations look like?
head(gapminder)
```

## ggplot grammar

ggplot doesn't work like a standard function in R in that you don't put every argument inside the function. Instead, you first create your plot and add geom, stat, scale, theme, facet and labels using the "+" sign. The most basic functions are: 

* **geoms**: Tell R what you want to do with the data (what type of graph you want to create)
* **labs**: Used to label the axis and add a title to your plot
* **themes**: Used to customize the non-data components of your plots (titles, labels, fonts, background, gridlines, and legends)

The ggplot() function is used to begin the plotting process. In and of itself, it doesn't plot anything - it simply provides the basic structure of the graph. You can think of it as the "skeleton" of your graph. 

```{r}
# ggplot requires data and mappings. In this case we want to understand the relationship between gdp and life expectancy
ggplot(data = gapminder,
       mapping = aes(x = gdpPercap, y = lifeExp))
```

We then add layers on top. Note that the arguments used inside aes() will be passed to the different geoms that we may use. 

## geom_point

Use geom_point to create a scatterplot

```{r, eval = FALSE}
# Adding a geom_point to create a scatterplot
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()

# We can also add colours, sizes and facets
# Here the col argument is *inside* the aes() for geom_point, so the color
# will be determined by the values of a particular variable
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent))

# If I simply want a color for all points, I can specify the color outside of aes()
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(col = "blue")

# I can use facet_grid() to create multiple panels based on some variable
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent)) +
  facet_grid(~continent)

# With horizontal panels instead
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent)) +
  facet_grid(continent~.)

# I can modify the size based on population size 
# Again, if I wanted to change the default size for all points, 
# I could use the size argument outside aes()
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop))

# And all sorts of other things
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(fill = continent, size = pop, alpha = pop), pch = 21) +
  facet_grid(continent~.)

# As an side...You can even make CATTER PLOTS!!!
install_github("Gibbsdavidl/CatterPlots")
library(CatterPlots)

CatterPlots::multicat(xs= runif(12, 0, 1), ys=runif(12, 0, 1), cat=c(1, 2, 3, 4, 5), catcolor='#000000FF')
```

ggplot is very powerful. I encourage you all to take a look at the ggplot lessons on datacamp.

## geom_histogram 

Use geom_histogram to create a histogram

```{r, eval = FALSE}
# lets say we want a histogram of gdpPercap
ggplot(data = gapminder, mapping = aes(x = gdpPercap)) +
  geom_histogram()

# Modifying the binwidth
ggplot(data = gapminder, mapping = aes(x = gdpPercap)) +
  geom_histogram(binwidth = 100)
```

This is a bit hard to interpret, so let's restrict to just countries with less than 45000 gdp per capita, and add a facet grid.

```{r message = FALSE}
filter(gapminder, gdpPercap <= 45000) %>%
  ggplot(mapping = aes(x = gdpPercap)) +
  geom_histogram() +
  facet_grid(~continent) +
  scale_x_discrete(limits = c(0, 20000, 40000))
```

## geom_density
Similar to geom_histogram, but very useful if you want to compare distributions in the same plot

```{r}
# Using filter() and a pipe, so I don't have to specify the data argument
filter(gapminder, continent %in% c("Africa", "Asia")) %>% 
  ggplot(mapping = aes(x = gdpPercap, fill = continent)) +
  geom_density(alpha = 0.7)

# Aplying a log transformation.
# What can yousay about these distributions?
filter(gapminder, continent %in% c("Africa", "Asia")) %>% 
  ggplot(mapping = aes(x = log(gdpPercap), fill = continent)) +
  geom_density(alpha = 0.7)
```

## geom_boxplot

Use geom_boxplot to create a boxplot

```{r eval=FALSE}
ggplot(data = gapminder, mapping = aes(y = lifeExp)) +
  geom_boxplot()
```

```{r}
# We need to add an x value
ggplot(data = gapminder, mapping = aes(x = continent, y = gdpPercap)) +
  geom_boxplot()
```

## Exercise

Again, we have the same problem - can we rerun this chart on a single year of data with countries with gdp per capitas lower than 45000 and choose only one year?


## Labels

Simply use the labs function to specify the title of the graph and the labels of the axis

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "GDP per Capita and life expectancy",
       subtitle = "hello yes there is more information here",
       x = "GDP per Capita",
       y = "Life expectancy",
       caption = "even more information here!")
```

## Themes

Themes allow you to customize the non-data components of your plots. E.g. theme_bw() produces a nice black and white graph:

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "GDP per Capita and life expectancy",
       x = "GDP per Capita",
       y = "Life expectancy") +
  theme_bw()
```

Some themes can also help you format your graph to fit in a particular newspaper/magazine:

- theme_stata()
- theme_wsj()
- theme_economist()

These are from the ggthemes package. 

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  theme_economist()
```

Here is a list of functions if you want to modify elements of a theme (e.g. size and position of the title, axis and legend): 
https://ggplot2.tidyverse.org/reference/theme.html 

Example:

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  theme_economist() +
  theme(legend.text = element_text(size = 7),
        legend.position = "right",
        legend.direction = "vertical")
```

You can even create a custom theme. 

What if we want to get rid of the legend? 

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  guides(col = FALSE,
         size = FALSE)
```

What about the axes?

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  scale_x_continuous(limits = c(0, 120000),
                     breaks = seq(from = 0, to = 120000, by = 10000))
```

What if I want to change the colors created using aes?

```{r}
# You can manually specify the colors
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  scale_x_continuous(limits = c(0, 120000),
                     breaks = seq(from = 0, to = 120000, by = 10000)) +
  scale_color_manual(name = "Continent", # Name of the legend
                     values = c("blue", "red", "black", "grey", "yellow"),
                     labels = c("Continent 1", "Continent 2", "another one...", "continent 4", "last continent"))

# Or use a pre-existing palette. I quite like palettes from the viridis package
# You can manually specify the colors
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  scale_x_continuous(limits = c(0, 120000),
                     breaks = seq(from = 0, to = 120000, by = 10000)) +
  scale_color_manual(values = viridis(5))

# all viridis does is poduce a vector of HEX codes 
viridis(5)
```

## Interactive plots

We can use the plotly package to create interactive plots. 
```{r}
# Everything inside aes() will be used to create a nice panel 
# that gives information about each observation. 
# I add label = country to also get the country name
gdp_plot <- filter(gapminder, year == 2007) %>% 
  ggplot(mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(col = continent, size = pop, label = country)) +
  facet_grid(continent~.) +
  labs(title = "Do rich people live longer?",
       x = "GDP per capita",
       y = "Life expectancy") +
  scale_x_continuous(limits = c(0, 120000),
                     breaks = seq(from = 0, to = 120000, by = 10000)) +
  scale_color_manual(values = viridis(5))

ggplotly(gdp_plot)
```

## Neat trick to highlight certain observations 
```{r}
gapminder <- gapminder %>% 
  mutate(can_usa = ifelse(country %in% c("Canada", "United States"), 1, 0) %>% factor())

ggplot(gapminder, aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(col = can_usa,
                 alpha = ifelse(can_usa == "1", 1, 0.5))) +
  scale_color_manual(values = c("0" = "black", "1" = "red")) +
  guides(col = FALSE,
         alpha = FALSE)
```

# Exercise

I want to see the evolution of life expectancy over time in each of the continent. 

```{r}

```



## ggplot tips and tricks
  
We are going to use the Prestige dataset from the car package. The dataset was constructed using 1971 census data from Canada. Each observation of the dataset (n = 102) is an occupation (e.g. biologists, secretaries, athletes, etc.) and the variables are: the average `education` and `income` for the respondents with those occupations, the percentage of `women` respondents, the `prestige` of the occupation (Pineo-Porter prestige score: social survey from the 1960s), and `type`of occupation (Blue collar -bc-, Professional -prof-, and White Collar -wc-).

```{r, eval = FALSE}
library(car)
glimpse(Prestige)
# qplot is a shortcut for ggplot that more closely resembles plot
qplot(x = income, geom="density", data = Prestige)
```

This chart shows an uneven distribution, but how uneven? You might want to show it next to what a normal distribution would be. We can visualize it compared to a normal distribution using the stat_function geom from ggplot.

```{r}
# first we get the mean and sd of income so we can produce a dnorm
sd_income <- sd(Prestige$income)
mean_income <- mean(Prestige$income)
# next we do our basic qplot but this time we add (just like ggplot) a stat_function, which we pass a function to (dnorm) and then args of the dnorm function as a list
qplot(x = income, geom="density", data = Prestige, colour = I("blue")) +
  stat_function(col = "red",
                fun = dnorm, 
                args = list(mean = mean_income, sd = sd_income))
# args: list of the arguments that this function (here: dnorm) takes (mean, and sd). 
```

```{r}
# this can of course be done in ggplot as well (with many more options for customization!)
ggplot(data = Prestige, aes(x = income)) +
  geom_line(stat = "density", col = "blue", size = 1) + 
  stat_function(col = "red",
                size = 1,
                fun = dnorm, 
                args = list(mean = mean_income, sd = sd_income)) +
  theme(legend.position="bottom")
# You can also use geom_density rather than geom_line
ggplot(data = Prestige, aes(x = income)) +
  geom_density(col = "blue", size = 1) + 
  stat_function(col = "red",
                size = 1,
                fun = dnorm, 
                args = list(mean = mean_income, sd = sd_income)) +
  theme(legend.position="bottom")
```

Note that qplot can work with vectors while ggplot needs a data frame. If you are trying to show a normal distribution, you can use something like:

```{r}
ggplot(data.frame(values = c(-3, 3)), aes(x = values)) +
  stat_function(fun = dnorm)
# default: mean = 0, sd = 1 
# You can technically use any type of function you want with stat_function
functionx <- function(x){
  x^3
}
ggplot(data.frame(values = c(-3, 3)), aes(x = values)) +
  stat_function(fun = functionx)
```
